---
title: "Exploratory Analysis"
format: html
editor: visual
---

## Introduction

The purpose of this document is to showcase our exploratory data analysis surrounding factors that may contribute to diabetes (or lack thereof). Our data comes from Behavioral Risk Factor Surveillance System dataset on [Kaggle](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/) where the organization phoned over 400,000 individuals in 2015 asking about the presence of diabetes and other factors.

We aim to predict diabetes based on these variables and understand our most important/impactful variables.

## Libraries & Reading Data

Below we will read in our libraries and our diabetes data set. As we can see in the output, some of our "numeric" variables shouldn't be numeric as they are categorical variables.

```{r}
library(tidyverse)

raw_data <- read_csv(file='diabetes_data.csv',show_col_types=FALSE)
head(raw_data)
```

## Data Cleaning

As stated, a lot of our data is labeled as a numeric type but should be categorical. Lets change all those for factors, and for everything that is not binary we will provide labels explicitly.

```{r}
data <- raw_data %>%
  mutate(Diabetes_binary = as.factor(Diabetes_binary),
         HighBP = as.factor(HighBP),
         HighChol = as.factor(HighChol),
         CholCheck = as.factor(CholCheck),
         Smoker = as.factor(Smoker),
         Stroke = as.factor(Stroke),
         HeartDiseaseorAttack = as.factor(HeartDiseaseorAttack),
         PhysActivity = as.factor(PhysActivity),
         Fruits = as.factor(Fruits),
         Veggies = as.factor(Veggies),
         HvyAlcoholConsump = as.factor(HvyAlcoholConsump),
         AnyHealthcare = as.factor(AnyHealthcare),
         NoDocbcCost = as.factor(NoDocbcCost),
         GenHlth = factor(GenHlth,levels=seq(1:5),labels=c('Excellent','Very Good','Good','Fair','Poor')),
         DiffWalk = as.factor(DiffWalk),
         Sex = as.factor(Sex),
         Age = factor(Age,levels=seq(1:13),labels=c('18-24','25-29','30-34','35-39','40-44','45-49'
                                            ,'50-54','55-59','60-64','65-69','70-74','75-79','80+')
          ),
         Education = factor(Education,levels=seq(1:6),labels=c('Never attended school or only kindergarten','Grades 1-8','Grades 9-11','Grade 12 or GED','College 1-3 years','College 4+ years')),
         Income = factor(Income,levels=seq(1:8),labels=c('<$10k','<$15k','<$20k','<$25k','<$35k','<$50k','<$75k','$75k+'))
    )

head(data)
```

Now that we've converted some things to factors, lets check for nulls (NA). There are none! Great news, since we know that not accounting for nulls can hurt our data quality and modeling.

```{r}
colSums(is.na(data))
```

## Data Summaries

Exploring our data is important. In this section, we'll take a look at our categorical and numeric variables and trends with respect to our outcome (presence of diabetes). This will include tables and plots.

### Categorical

Starting with some contingency tables, lets look at two-way contingency tables. Based on what I know, I think that our `Age` variable (age brackets), `HighBP` (high blood pressure), and `Highhol` (high cholesterol) may provide some insights.

From the below outputs we see that

1.  It is difficult to observe trends in Age bracket and the presence of Diabetes, but we can roughly tell the distribution of our Age brackets. We'll explore this further in graphs.
2.  The majority of those that have diabetes have High Blood Pressure
3.  The majority of those that have diabetes have High Cholesterol
4.  At surface level, Sex seems to be of equal sample proportions for those with and without diabetes

```{r}
data %>%
  group_by(Age,Diabetes_binary) %>%
  summarize(count = n())

data %>%
  group_by(Diabetes_binary,HighBP) %>%
  summarize(count = n())

data %>%
  group_by(Diabetes_binary,HighChol) %>%
  summarize(count = n())

data %>%
  group_by(Diabetes_binary,Sex) %>%
  summarize(count = n())

```

Now for some three-way contingency tables. From the above, we saw that people with diabetes exhibited High Blood Pressure OR High Cholesterol. We want to see if that can also be an AND statement. Furthermore, I'm curious on if being a smoker and an assessment of general health has an impact.

From the below outputs we observe...

1.  High Cholesterol AND High Blood Pressure is often seen in those with diabetes
2.  Those without diabetes exhibit `Very Good` general health (mode) with perhaps a slight penalty for being a smoker.
3.  Those with diabetes seem to have slightly lower general health of `Good` with a slight shift in distribution towards `Fair` if they smoke

```{r}
data %>%
  group_by(Diabetes_binary,HighChol,HighBP) %>%
  summarize(count = n())

data %>%
  group_by(Diabetes_binary,Smoker,GenHlth) %>%
  summarize(count = n())
```

Now for some graphs! To start out, we made a statement earlier that it was difficult to infer anything from our Age and Diabetes contingency table earlier. Lets visualize this graphically. We see in the below that our distribution is left skewed for our respondents and it seems that there is little to know impact of age on the precense of diabetes.

```{r}
ggplot(data %>% 
         mutate(Diabetes_binary = factor(Diabetes_binary,levels=c(0,1),labels=c('No Diabetes','Has Diabetes'))),
       aes(x=Age,fill=Diabetes_binary)
       ) +
  geom_bar() +
  scale_fill_discrete("Diabetes") +
  labs(y = 'Frequency',title='Age Distribution colored by Diabetes')
  
```

Similar to Age, lets create similar bar graphs for levels of income and education. It is our hypothesis that there is no connection or obvious patterns here. Lets see what it looks like at first glance

Our hypothesis appears correct.

```{r}
ggplot(data %>% 
         mutate(Diabetes_binary = factor(Diabetes_binary,levels=c(0,1),labels=c('No Diabetes','Has Diabetes'))),
       aes(x=Income,fill=Diabetes_binary)
       ) +
  geom_bar() +
  scale_fill_discrete("Diabetes") +
  labs(y = 'Frequency',title='Income Distribution colored by Diabetes')

ggplot(data %>% 
         mutate(Diabetes_binary = factor(Diabetes_binary,levels=c(0,1),labels=c('No Diabetes','Has Diabetes'))),
       aes(x=Education,fill=Diabetes_binary)
       ) +
  geom_bar() +
  scale_fill_discrete("Diabetes") +
  labs(y = 'Frequency',title='Income Distribution colored by Diabetes')
```

Maybe we can find some patterns using faceting. Lets build a series of bar graphs with `GenHealth` on our X-Axis colored by the presence of diabetes. Lets facet this by our age brackets to see if there's a pattern.

The output isn't super friendly, but because we know our General Health is a factor type the axis goes from left-to-right in the following order: `Excellent`,`Very Good`,`Good`,`Fair`,`Poor`

Looking at General Health alone, it seems there's a slight right skew that becomes more normal with age. Its difficult to tell, but there may be some suggestion that your general health leads to an increased risk of diabetes due to the eye-level proportions.

```{r}
ggplot(data %>% 
         mutate(Diabetes_binary = factor(Diabetes_binary,levels=c(0,1),labels=c('No Diabetes','Has Diabetes'))),
       aes(x=GenHlth,fill=Diabetes_binary)
       ) +
  geom_bar() +
  scale_fill_discrete("Diabetes") +
  labs(x='General Health',y = 'Frequency',title='General Health by Age Bracket') +
  facet_wrap(~Age)


```

### Numeric

We have a lot less numeric variables here to analyze, but there still may be some important details in our summaries. Lets look at our distributions and measures of spread and central tendency.

For starters, lets look at standalone variables. Lets look at some stats for `BMI`,`MentHlth`,`PhysHlth`.

Below are some observations:

1.  Our mean and median BMI values are between 27-29 with a standard deviation of 6. With a slight left skew, both values fall into the "Overweight" category.
2.  The median number of days (over past 30 days) our participants expressed having some mental health problems is 0. This is a bit odd and may showcase some bias in how our participants evaluate themselves.
3.  Yikes! The median number of days (over past 30 days) our participants expressed participating in physical activities is also 0! Not a very active bunch!

```{r}
data %>%
  summarize("mean" = mean(BMI),
            "median" = median(BMI),
            "var" = var(BMI),
            "sd" = sd(BMI),
            "IQR" = IQR(BMI)
            )

data %>%
  summarize("mean" = mean(MentHlth),
            "median" = median(MentHlth),
            "var" = var(MentHlth),
            "sd" = sd(MentHlth),
            "IQR" = IQR(MentHlth)
            )

data %>%
  summarize("mean" = mean(PhysHlth),
            "median" = median(PhysHlth),
            "var" = var(PhysHlth),
            "sd" = sd(PhysHlth),
            "IQR" = IQR(PhysHlth)
            )

```

Lets go a step beyond one-variable summaries here. Mental Health and Physical Health seem that they may have their own problems. Fortunately for us, we have a BMI number. We would expect our measures of central tendency to be higher, on average, for those with diabetes.

Our hypothesis seems correct at first glance with our median BMI being 27 (Overweight) for those without diabetes and 31 (Obese) for those with diabetes. We'll explore this more in our graphs.

```{r}
data %>%
  select(Diabetes_binary,BMI) %>%
  group_by(Diabetes_binary) %>%
  summarize(across(everything(), .fns = list("mean" = mean,
                                       "median" = median,
                                       "var" = var,
                                       "sd" = sd,
                                       "IQR" = IQR), .names = "{.fn}_{.col}"))
```

Lets graph out the above relationship on a density chart. We can certainly see a shift to the right (higher BMI) for those with diabetes.

```{r}
bmi_plot <- ggplot(data %>% 
         mutate(Diabetes_binary = factor(Diabetes_binary,levels=c(0,1),labels=c('No Diabetes','Has Diabetes')),
                HighChol = factor(HighChol,levels=c(0,1),labels=c('Low/Average Cholesterol','High Cholesterol')),
                HighBP = factor(HighBP,levels=c(0,1),labels=c('Low/Average Blood Pressure','High Blood Pressure'))
                ),
       aes(x=BMI,fill=Diabetes_binary)
       ) +
  geom_density(alpha=0.5,kernel="gaussian") +
  scale_fill_discrete("Diabetes") +
  labs(x='BMI',y='Frequency')

bmi_plot +
  labs(title='BMI')
```

It would be interesting to see this faceted across our categorical `HighChol` and `HighBp` measures. We would expect to see some patterns.

It is difficult to tell in the graphs, but the graphs look similar across low and high levels of each categorical. There may be a slight shift to the right (BMI) in measures of central tendency across high levels of each categorical. We'll uncover more on the impact of this in our modeling!

```{r}
bmi_plot +
  labs(title = 'BMI Distribution Across Generic Cholesterol Levels') +
  facet_wrap(~HighChol)

bmi_plot +
  labs(title = 'BMI Distribution Across Generic Blood Pressure Levels') +
  facet_wrap(~HighBP)
```

From our previous analyses, I hypothesize that mental health and physical health have little to no influence, and its hard to think a box plot or violin plot will provide much information.

What about our correlation between our numeric variables.? It seems BMI and our numeric health metrics have little correlation, but we see low-to-moderate positive correlation between `MentHlth` and `PhysHlth`.

```{r}
correlations <- summary(correlation::correlation(data %>% select(BMI, MentHlth, PhysHlth)))

plot(correlations, show_data = "points")
```

## Summary

That concludes our exploratory data analysis. We've seen a lot of interesting trends or absence of trends. Generally, here are some highlights:

1.  BMI seems to be higher for those with diabetes
2.  Mental Health and Physical Health measures seem to have some correlation, but none that seem to impact BMI or classification of someone having diabetes.
3.  General Health may have some impact on developing diabetes.
4.  High Blood Pressure and/or High Cholesterol are relatively more common in those with diabetes than those without.
